Data Exploration coral maps: IUCN and AquaMaps

### 2016-03-27 data_explore/data_explore_coralmaps.Rmd
trying to get a coral map going

---
title: 'Data Exploration quad maps: IUCN and AquaMaps'
output: html_document
---

@@

@@

# Data files
#iucn spp cells WITHOUT extinct locations of IUCN subpopulations
iucn_spp_cells <- read_csv(file.path(dir_anx, 'explore_am_v_iucn/iucn_spp_cells_2015.csv'), 
                           col_types = 'cddddc', progress = TRUE) %>%
  as.data.frame() %>%
  filter(presence != 5) ### removes extinct subpopulations

am_spp_cells   <- read_csv(file.path(dir_anx, 'explore_am_v_iucn/am_spp_cells_2015.csv'), progress = TRUE)

### LOICZID cell ID numbers - used in plotting rasters
loiczid_raster_file  <- file.path(dir_git, 'shiny/data/loiczid_raster.grd')
loiczid_raster       <- raster(loiczid_raster_file)
names(loiczid_raster) <- 'loiczid'
```

```{r get_spp_list, echo = FALSE, warning = FALSE, message = FALSE, eval = TRUE}
### this file created in data_prep
spp_list <- read.csv(file.path(dir_data, 'spp_list_w_area_trimmed.csv'), stringsAsFactors = FALSE)

@@

Quadrant map and quad breakdown by taxa and risk

``` {r pct_overlap_vs_ratio_of_area, echo = FALSE, warning = FALSE, message = FALSE, eval = TRUE}

spp_list <- spp_list %>%
  mutate(area_ratio = am_area / iucn_area,
         area_ratio = ifelse(area_ratio > 1, 1 / area_ratio, area_ratio),
         area_ratio = area_ratio * 100)

### define 25%, 50%, and 75% quartiles for both x and y axes
area_ratio_quartiles <- quantile(spp_list$area_ratio, na.rm = TRUE)[2:4]
pct_overlap_quartiles <- quantile(spp_list$sm_perc, na.rm = TRUE)[2:4]

# spp_rando <- c("Lagocephalus lagocephalus",
#                "Makaira nigricans",
#                "Pentacheles laevis",
#                "Stereomastis suhmi",
#                "Stereomastis surda") 
# 
### truncate the species names for the representative species
# spp_list_label <- spp_list %>%
#   filter(sciname %in% spp_rando) %>%
#   separate(sciname, c('g', 's'), sep = ' ', remove = FALSE) %>%
#   mutate(sciname_label = paste(str_sub(g, 1, 1), '. ', s, sep = '')) %>%
#   dplyr::select(-g, -s)

### mongo plot time
scatter_arearatio_vs_overlap <- ggplot(spp_list %>% 
                                         transform(popn_category = reorder(popn_category, category_score)),
                                       aes(x = area_ratio, y = sm_perc, 
                                           color = popn_category,
                                           size = lg_area_pct)) + 
  ggtheme_plot + 
  ### color the quadrant backgrounds:
    annotate("rect", xmin = area_ratio_quartiles[2],  xmax = 100, 
                     ymin = pct_overlap_quartiles[2], ymax = 100, alpha = .3, fill= "green4")  + 
    annotate("rect", xmin = area_ratio_quartiles[2],  xmax =   0, 
                     ymin = pct_overlap_quartiles[2], ymax = 100, alpha = .3, fill= "greenyellow") + 
    annotate("rect", xmin = area_ratio_quartiles[2],  xmax = 100, 
                     ymin = pct_overlap_quartiles[2], ymax =   0, alpha = .3, fill= "orange") + 
    annotate("rect", xmin = area_ratio_quartiles[2],  xmax =   0, 
                     ymin = pct_overlap_quartiles[2], ymax =   0, alpha = .3, fill= "orangered2") + 
  geom_point(alpha = .6)

### Manage scales for color and size 
scatter_arearatio_vs_overlap <- scatter_arearatio_vs_overlap +
  scale_colour_manual(values = c('LC' = 'green4', 
                                 'NT' = 'greenyellow', 
                                 'VU' = 'yellow', 
                                 'EN' = 'orange', 
                                 'CR' = 'orangered2',
                                 'DD' = 'grey30')) + 
  scale_size_continuous(limits = c(0, 100),
                        breaks = c( 0,    25,    50,    75,    100), 
                        labels = c('0%', '25%', '50%', '75%', '100%')) + 
  scale_x_continuous(expand = c(0, 0), limits = c(0, 100)) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 100))

### here are median labels:
scatter_arearatio_vs_overlap <- scatter_arearatio_vs_overlap +
  annotate(geom = 'text',
           x = area_ratio_quartiles[2], y = 5,
           hjust = 0, vjust = 0,
           color = 'grey50', size = 2,
           fontface = 'bold.italic', angle = 90,
           label = sprintf('Median = %s%%', round(area_ratio_quartiles[2], 1))) +
  annotate(geom = 'text',
           x = 5, y = pct_overlap_quartiles[2],
           hjust = 0, vjust = 0,
           color = 'grey50', size = 2,
           fontface = 'bold.italic', angle = 0,
           label = sprintf('Median = %s%%', round(pct_overlap_quartiles[2], 1)))

### here are median lines:
# scatter_arearatio_vs_overlap <- scatter_arearatio_vs_overlap +
#   geom_vline(x = area_ratio_quartiles[2],
#              color = 'grey50', linetype = 'dashed') +
#   geom_hline(y = pct_overlap_quartiles[2],
#              color = 'grey50', linetype = 'dashed')

### here are 25% and 75% quantile lines:
# scatter_arearatio_vs_overlap <- scatter_arearatio_vs_overlap +
#   geom_vline(x = area_ratio_quartiles[c(1, 3)],
#              color = 'grey50', linetype = 'dashed') +
#   geom_hline(y = pct_overlap_quartiles[c(1, 3)],
#              color = 'grey50', linetype = 'dashed')

### here are individual species, identified and named on the plot:
# scatter_arearatio_vs_overlap <- scatter_arearatio_vs_overlap +
#   geom_point(data = spp_list_label, shape = 4, color = 'grey30', size = 4) + 
#   geom_text(data = spp_list_label, 
#             aes(label = sciname_label), 
#             color = 'grey30',
#             fontface = 'bold.italic',
#             size = 2,
#             hjust = .25,
#             vjust = 1.5)

scatter_arearatio_vs_overlap <- scatter_arearatio_vs_overlap +
  labs(x = bquote('Extent alignment '*alpha['ext']*' (%)'), 
       y = bquote('Distribution alignment '*alpha['dist']*' (%)'), 
       title = 'Distribution alignment vs Extent alignment',
       color = 'Risk', size = '% ocean range')

print(scatter_arearatio_vs_overlap)

``` {r plot_spp_from_quadrants, echo = FALSE, warning = FALSE, message = FALSE, eval = TRUE}
### define representative species from each quadrant, for over-plotting
spp_quads <- c('q1' = 'Kajikia albida', 
               'q2' = 'Conus episcopatus',
               'q3' = 'Conus magnificus', 
               'q4' = 'Acanthurus nigroris')
am_thresh <- 0

for (i in 1:length(spp_quads)) { # i = 1
  spp <- spp_list %>% filter(sciname == spp_quads[i])
  spp_quad <- names(spp_quads)[i]

  spp_plot <- plot_rangemap(spp)
  spp_plot <- spp_plot + 
    labs(title = element_blank())

>>>>

---
title: 'Data Exploration coral maps: IUCN and AquaMaps'
output: html_document
---

@@

``` {r define_fxns, echo = FALSE}
#################################################################=
### Species Map Function ###
# This function takes a dataframe of loiczid per species ID, and a target species ID 
### (as single or vector).  Return a df of cells ready to be rasterized.  
### If AquaMaps, returned df includes aquamaps probability; if IUCN, returned
### df includes proportional area.
get_iucn_spp_map <- function(sid, spp_cells) {

  spp_map <- spp_cells %>%
    filter(iucn_sid %in% sid)
  if(nrow(spp_map) == 0) {
    message('no IUCN cells found')
    spp_map <- spp_map %>%
      as.data.frame() %>%
      select(iucn_sid, loiczid) %>%
      mutate(area_iucn = NA)
  } else {
    spp_map <- spp_map %>%
      group_by(iucn_sid, loiczid) %>%        
      summarize(area_iucn = max(prop_area))
    ### The group_by() and summarize() are to collapse cells with 
    ### overlapped polygons, fixing the duplicate 'by' issue.
    ### Note: this drops presence, sciname, and subpop columns
  }

  return(spp_map %>% unique() %>% as.data.frame())
}

#################################################################=
### Map vs Depth function
### uses ggplot to create a formatted map of species ranges, including
### Aquamaps, IUCN, and overlapped ranges; returns the plot object.
plot_depthmap <- function(spp) {
  map_iucn <- get_spp_map(spp$iucn_sid, iucn_spp_cells) %>%
    mutate(iucn_pres = ifelse(area_iucn > 0, 1, 0)) %>%
    select(-area_iucn, -iucn_sid) %>%
    unique()

  r_iucn_spp <- subs(loiczid_raster, 
                     map_iucn[ , c('loiczid', 'iucn_pres')], 
                     by = 'loiczid', 
                     which = 'iucn_pres', 
                     subsWithNA = TRUE)

  spp_pts <- as.data.frame(rasterToPoints(r_iucn_spp)) %>%
    mutate(presence = 'IUCN')

  depth_poly_df <- rgdal::readOGR(dsn = 'coralmaps', layer = 'bath_200m') %>%
    fortify() %>%
    mutate(x = long, y = lat)

  spp_plot <- ggplot(spp_pts, aes(x = x, y = y)) +
    ggtheme_map + 
    geom_raster(aes(fill = presence), alpha = .8) +
    scale_fill_manual(values = c('IUCN' = '#d95f02')) +
    geom_polygon(data = depth_poly_df, color = 'black', fill = 'yellow', alpha = .8) +
    borders('world', color='gray40', fill='gray45', size = .1) +  # create a layer of borders
    scale_x_continuous(breaks = seq(-180, 180, by = 30), expand = c(0, 2)) +
    scale_y_continuous(breaks = seq( -90,  90, by = 30), expand = c(0, 2))
}

@@

spp_corals <- 'Oculina varicosa'

# Data files
#iucn spp cells WITHOUT extinct locations of IUCN subpopulations
iucn_coral_cells <- read_csv(file.path(dir_anx, 'explore_am_v_iucn/iucn_spp_cells_2015.csv'), 
                           col_types = 'cddddc', progress = TRUE) %>%
  as.data.frame() %>%
  filter(presence != 5) %>% ### removes extinct subpopulations
  filter(sciname %in% spp_corals)

### LOICZID cell ID numbers - used in plotting rasters
loiczid_raster_file  <- file.path(dir_git, 'shiny/data/loiczid_raster.grd')
loiczid_raster       <- raster(loiczid_raster_file)
names(loiczid_raster) <- 'loiczid'

### this file created in data_prep
spp_list_corals <- read.csv(file.path(dir_data, 'spp_list_w_area_trimmed.csv'), stringsAsFactors = FALSE) %>%
  filter(sciname %in% spp_corals)

@@

``` {r plot_coral_spp_at_depth, echo = FALSE, warning = FALSE, message = FALSE, eval = TRUE}
### define representative species from each quadrant, for over-plotting

  spp <- spp_list_corals

  spp_plot <- plot_depthmap(spp)
